Carrying out Latent Profile Analysis in Accordance With Best Practices: A Tutorial Using the tidyLPA R package

### 2020-01-16 paper.rmd
update

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Enter author note here.

abstract: |

keywords          : "Latent Profile Analysis, mixture models, finite mixture models, tutorial, R, MPlus, mclust"
wordcount         : ""

@@

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

@@

```{r, include = FALSE}
library(simstudy)

@@

```{r, echo - FALSE}
library(tidyverse)
ggplot(dx, aes(x = xMix)) +
  geom_density() +
  papaja::theme_apa() +
  ylim(0, .10)
```

@@

```{r, echo = FALSE}
dx %>% 
  gather(key, val, x1:x2) %>% 
  ggplot(aes(x = val, fill = key)) +
  geom_density(alpha = .3) +
  papaja::theme_apa() +
  ylim(0, .10)
```

@@

<!-- This is where we need to use an example from our own data -->

@@

Here is a brief example using the built-in `pisaUSA15` data set and variables
for broad interest, enjoyment, and self-efficacy. Note that we first type the
name of the data frame, followed by the unquoted names of the variables used to
create the profiles. We also specify the number of profiles and the model. See
`?estimate_profiles` for more details.

@@

``` {r}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    estimate_profiles(3)
```

@@

``` {r, eval = FALSE}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    estimate_profiles(3, package = "MplusAutomation")
```

@@

``` {r}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    scale() %>%
    estimate_profiles(3) %>% 
    plot_profiles()
```

@@

``` {r, eval = F}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    estimate_profiles(3, 
                      variances = "varying",
                      covariances = "varying")
```

@@

``` {r, eval = FALSE}
pisaUSA15[1:100, ] %>%
    select(broad_interest, enjoyment, self_efficacy) %>%
    single_imputation() %>%
    estimate_profiles(1:3, 
                      variances = c("equal", "varying"),
                      covariances = c("zero", "varying")) %>%
    compare_solutions(statistics = c("AIC", "BIC"))
```

>>>>

abstract: 
keywords          : "Latent Profile Analysis, mixture models, finite mixture models, tutorial, R, MPlus, mclust"
wordcount         : ""

@@

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
always_allow_html: true
---

```{r setup, include = FALSE}
library(papaja)
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
library(tidyverse)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed, cache = TRUE)
```

@@

```{r, include = FALSE, eval = FALSE}
library(simstudy)

@@

```{r, echo = FALSE, eval = FALSE}

ggplot(dx, aes(x = xMix)) +
  geom_density() +
  papaja::theme_apa() +
  ylim(0, .10)
```

@@

```{r, echo = FALSE, eval = FALSE}
dx %>% 
  gather(key, val, x1:x2) %>% 
  ggplot(aes(x = val, fill = key)) +
  geom_density(alpha = .3) +
  papaja::theme_apa() +
  ylim(0, .10)
```

@@

Here, we use a simulated dataset, based on Curry et al., 2019, contains data on moral relevance and judgment across the seven domains of the *Morality As Cooperation* scale.

@@

@@

``` {r}
curry_mac %>% 
  select(KinshipR:PossessionR, 
         KinshipJ:PossessionJ) %>% 
  estimate_profiles(3)
```

@@

``` {r, eval = FALSE}
curry_mac %>% 
  select(KinshipR:PossessionR, 
         KinshipJ:PossessionJ) %>% 
  estimate_profiles(3, package = "MplusAutomation")
```

@@

```{r}
curry_mac %>% 
  select(KinshipR:PossessionR, 
         KinshipJ:PossessionJ) %>% 
  estimate_profiles(3) %>% 
  plot_profiles()
```

@@

``` {r, eval = F}
curry_mac %>% 
  select(KinshipR:PossessionR, 
         KinshipJ:PossessionJ) %>% 
  estimate_profiles(3, 
                    variances = "varying",
                    covariances = "varying")
```

@@

``` {r, compare-mclust, eval = FALSE}
curry_mac %>% 
  select(KinshipR:PossessionR, 
         KinshipJ:PossessionJ) %>% 
  estimate_profiles(models = c(1, 2, 3, 6),
                    n_profiles = 1:9) %>%
  compare_solutions(statistics = c("AIC", "BIC"))
```

We can also run the same process using mplus:

``` {r, compare-mplus, eval = FALSE}
curry_mac %>% 
  select(KinshipR:PossessionR, 
         KinshipJ:PossessionJ) %>% 
  estimate_profiles(models = c(1, 2, 3, 6),
                    n_profiles = 1:9, package = "mplus") %>%
  compare_solutions(statistics = c("AIC", "BIC"))
```

### 2020-01-16 paper.tex
update

\authornote{Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

Enter author note here.

@@

\includegraphics{paper_files/figure-latex/echo - FALSE-1.pdf}

@@

\includegraphics{paper_files/figure-latex/unnamed-chunk-2-1.pdf}

@@

@@

Here is a brief example using the built-in \texttt{pisaUSA15} data set and variables
for broad interest, enjoyment, and self-efficacy. Note that we first type the
name of the data frame, followed by the unquoted names of the variables used to
create the profiles. We also specify the number of profiles and the model. See
\texttt{?estimate\_profiles} for more details.

@@

\begin{verbatim}
## tidyLPA analysis using mclust: 
## 
##  Model Classes AIC    BIC    Entropy prob_min prob_max n_min n_max BLRT_p
##  1     3       629.10 665.57 0.82    0.83     0.96     0.03  0.70  0.01
\end{verbatim}

@@

\includegraphics{paper_files/figure-latex/unnamed-chunk-7-1.pdf}

@@

@@

\includegraphics{paper_files/figure-latex/unnamed-chunk-13-1.pdf}

@@

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-R-tidyLPA}{}%
Rosenberg, J. M., Beymer, P. N., Anderson, D. J., Lissa, C. van, \& Schmidt, J. A. (2018). TidyLPA: An r package to easily carry out latent profile analysis (lpa) using open-source or commercial software. \emph{Journal of Open Source Software}, \emph{3}(30), 978. doi:\href{https://doi.org/10.21105/joss.00978}{10.21105/joss.00978}

>>>>

\authornote{

@@

@@

@@

Here, we use a simulated dataset, based on Curry et al., 2019, contains data on moral relevance and judgment across the seven domains of the \emph{Morality As Cooperation} scale.

@@

@@

\begin{verbatim}
## tidyLPA analysis using mclust: 
## 
##  Model Classes AIC       BIC       Entropy prob_min prob_max n_min n_max BLRT_p
##  1     3       158594.55 158898.38 0.82    0.91     0.92     0.15  0.53  0.01
\end{verbatim}

@@

\includegraphics{paper_files/figure-latex/unnamed-chunk-8-1.pdf}

@@

We can also run the same process using mplus:

@@

\hypertarget{htmlwidget-a6cf205f9fe062a59d12}{}

@@

\hypertarget{implications}{%
\section{Implications}\label{implications}}

A step away from the tutorial side of this work, we wish to note some implications
for the developers of statistical software for psychological scientists.

First, we note that developing a tool that emphasized both rigor and access has
been, so far, well-received. This suggests that there is an audience (or a
market) for tools to carry out sophisticated tools that may be otherwise
out-of-reach. Tools do not need to be hard to use, a point made by developers
of, for example, tools to carry out highly-sophisticated Bayesian methods using
the most cutting-edge tools (Bürkner \& others, 2017, @makowski2019). We think
that tidyLPA speaks to this, too.

Another implication concerns how we developed tidyLPA. We used a modern software
development workflow, including git/GitHub, a suite of tests, and continuous
integration tools. We also note that we emphasized communicating that we were
developing the software; when it was updated; and how we made major updates.

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-burkner2017}{}%
Bürkner, P.-C., \& others. (2017). Brms: An r package for bayesian multilevel models using stan. \emph{Journal of Statistical Software}, \emph{80}(1), 1--28.

\leavevmode\hypertarget{ref-makowski2019}{}%
Makowski, D., Ben-Shachar, M. S., Chen, S., \& Lüdecke, D. (2019). Indices of effect existence and significance in the bayesian framework. \emph{Frontiers in Psychology}, \emph{10}, 2767.

\leavevmode\hypertarget{ref-R-tidyLPA}{}%
Rosenberg, J. M., Beymer, P. N., Anderson, D. J., Lissa, C. van, \& Schmidt, J. A. (2018). TidyLPA: An r package to easily carry out latent profile analysis (lpa) using open-source or commercial software. \emph{Journal of Open Source Software}, \emph{3}(30), 978. doi:\href{https://doi.org/10.21105/joss.00978}{10.21105/joss.00978}
